---
title: "Pharmacokinetic Compartmental Modeling Report"
subtitle: "One-Compartment Oral Absorption Model — Theophylline Dataset"
author: "PKPD Analysis Project"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

```{css, echo=FALSE}
body {
  font-family: 'Segoe UI', system-ui, -apple-system, sans-serif;
  font-size: 15px;
  color: #2c2c3e;
  line-height: 1.7;
}
h1, h2, h3 { color: #1a1a2e; font-weight: 700; }
.title { color: #1a1a2e; }
.subtitle { color: #5555aa; }
table { font-size: 13px; }
.highlight { background: #f0f4ff; border-left: 4px solid #2196F3; padding: 10px 15px; border-radius: 4px; margin: 10px 0; }
code { color: #d63384; background: #fff0f5; padding: 2px 5px; border-radius: 3px; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  dpi = 120
)

library(nlme)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(patchwork)
```

---

# Executive Summary {-}

::: {.highlight}
This report presents a **one-compartment pharmacokinetic model** fitted to the classic **Theophylline dataset** using nonlinear mixed-effects (NLME) methodology. Theophylline is a bronchodilator with a narrow therapeutic index, making precise PK characterization clinically important.

**Key findings:** The one-compartment model with first-order absorption and elimination provided excellent fits. Population half-life was approximately **7.9 hours**, and clearance showed moderate inter-individual variability.
:::

---

# Background & Objectives

## Clinical Context

Theophylline is a methylxanthine used in the management of asthma and COPD. Its narrow therapeutic window (5–15 mg/L) necessitates therapeutic drug monitoring, making robust PK characterization essential.

## Dataset Description

The **Theophylline dataset** (Boeckmann et al., 1994) contains:

- **12 subjects** (healthy volunteers)
- Single oral dose of **theophylline** (dose range: ~3.1–5.9 mg/kg)
- **10–11 plasma samples per subject** over 25 hours
- Weight-adjusted dosing

## Objectives

1. Estimate population PK parameters (Ka, CL, V) using NLME modeling  
2. Quantify inter-individual variability (IIV)  
3. Assess model adequacy via diagnostic plots  
4. Report non-compartmental analysis (NCA) metrics

---

# Data Exploration

```{r load-data}
# Load processed data
pk_data        <- readRDS("../data/pk_data_processed.rds")
pk_summary     <- readRDS("../data/pk_summary.rds")
param_table    <- readRDS("../data/param_table.rds")
pk_nlme        <- readRDS("../data/pk_nlme_model.rds")
pop_pred_smooth <- readRDS("../data/pop_pred_smooth.rds")

cat(sprintf("Dataset: %d subjects, %d total observations\n",
    n_distinct(pk_data$subject), nrow(pk_data)))
```

## Dataset Overview

```{r data-table}
pk_data %>%
  group_by(subject) %>%
  slice(1) %>%
  select(subject, wt, dose, dose_mg) %>%
  rename(Subject = subject, `Weight (kg)` = wt,
         `Dose (mg/kg)` = dose, `Total Dose (mg)` = dose_mg) %>%
  kable(caption = "Study Design: Dose and Subject Information") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

## Concentration-Time Profiles

```{r conc-time-plot, fig.height=6}
subject_colors <- colorRampPalette(
  c("#2196F3","#E91E63","#4CAF50","#FF9800","#9C27B0","#00BCD4",
    "#F44336","#795548","#607D8B","#FFEB3B","#3F51B5","#FF5722"))(12)

pk_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13, color = "#1a1a2e"),
    plot.subtitle = element_text(color = "#555577", size = 10),
    axis.title    = element_text(face = "bold", color = "#333355"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#eeeeee"),
    strip.text    = element_text(face = "bold"),
    legend.position = "right"
  )

ggplot(pk_data, aes(x = time, y = conc, color = subject, group = subject)) +
  geom_line(alpha = 0.7, linewidth = 0.9) +
  geom_point(size = 2.5, alpha = 0.9) +
  geom_line(data = pop_pred_smooth, aes(x = time, y = conc),
            inherit.aes = FALSE, color = "#1a1a2e", linewidth = 1.4,
            linetype = "dashed") +
  scale_color_manual(values = subject_colors) +
  labs(title = "Observed Concentration–Time Profiles",
       subtitle = "Individual curves (colored) | Population prediction (dashed)",
       x = "Time (h)", y = "Theophylline Concentration (mg/L)", color = "Subject") +
  pk_theme
```

---

# Non-Compartmental Analysis (NCA)

```{r nca-table}
pk_summary %>%
  mutate(across(where(is.numeric), ~round(.x, 2))) %>%
  rename(Subject = subject, `Weight (kg)` = weight_kg,
         `Dose (mg/kg)` = dose_mg_kg, `Total Dose (mg)` = total_dose,
         `Cmax (mg/L)` = Cmax, `Tmax (h)` = Tmax,
         `AUC₀₋ₜ (mg·h/L)` = AUC_trap) %>%
  kable(caption = "NCA Parameters by Subject (Trapezoidal Method)") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE)
```

```{r nca-boxplot, fig.height=4}
pk_summary %>%
  pivot_longer(cols = c(Cmax, AUC_trap, Tmax), names_to = "param", values_to = "value") %>%
  mutate(param = factor(param,
    levels = c("Cmax","AUC_trap","Tmax"),
    labels = c("Cmax (mg/L)","AUC (mg·h/L)","Tmax (h)"))) %>%
  ggplot(aes(x = param, y = value, fill = param)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.7, color = "#333") +
  scale_fill_manual(values = c("#2196F3","#E91E63","#4CAF50"), guide = "none") +
  labs(title = "Distribution of NCA Parameters Across Subjects",
       x = NULL, y = "Value") +
  pk_theme
```

---

# Compartmental PK Modeling

## Model Specification

The **one-compartment model with first-order oral absorption** was used:

$$C(t) = \frac{F \cdot D \cdot k_a}{V(k_a - k_e)} \left[ e^{-k_e t} - e^{-k_a t} \right]$$

Where:

| Parameter | Symbol | Units | Description |
|-----------|--------|-------|-------------|
| Absorption rate | $k_a$ | h⁻¹ | First-order absorption rate constant |
| Elimination rate | $k_e$ | h⁻¹ | First-order elimination rate constant |
| Clearance | $CL$ | L/h/kg | Apparent oral clearance (CL = ke × V) |
| Volume | $V$ | L/kg | Apparent volume of distribution |

> **NLME Framework:** Random effects on all three parameters (lKe, lKa, lCl) estimated using `nlme::nlme()` with diagonal random effects covariance structure.

## Population Parameter Estimates

```{r param-table}
param_table %>%
  kable(caption = "Population PK Parameter Estimates") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  column_spec(2, bold = TRUE, color = "#2196F3")
```

## Model Fit Summary

```{r model-summary}
summary(pk_nlme)
```

---

# Model Diagnostics

## Goodness-of-Fit Plots

```{r gof-plots, fig.height=10}
p_ipred <- ggplot(pk_data, aes(x = pred_ind, y = conc, color = subject)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#888", linewidth = 0.8) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_manual(values = subject_colors, guide = "none") +
  labs(title = "Individual Predicted vs Observed",
       x = "Individual Predicted (mg/L)", y = "Observed (mg/L)") + pk_theme

p_ppred <- ggplot(pk_data, aes(x = pred_pop, y = conc, color = subject)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#888", linewidth = 0.8) +
  geom_smooth(method = "loess", se = TRUE, aes(group = 1),
              color = "#E91E63", fill = "#E91E63", alpha = 0.15, linewidth = 1) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_manual(values = subject_colors, guide = "none") +
  labs(title = "Population Predicted vs Observed",
       x = "Population Predicted (mg/L)", y = "Observed (mg/L)") + pk_theme

p_restime <- ggplot(pk_data, aes(x = time, y = resid, color = subject)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#888", linewidth = 0.8) +
  geom_hline(yintercept = c(-2, 2), linetype = "dotted", color = "#cc4444", linewidth = 0.6) +
  geom_smooth(method = "loess", se = FALSE, aes(group = 1), color = "#FF9800", linewidth = 1) +
  geom_point(size = 2.2, alpha = 0.85) +
  scale_color_manual(values = subject_colors, guide = "none") +
  labs(title = "Residuals vs Time", x = "Time (h)", y = "Normalized Residuals") + pk_theme

p_respred <- ggplot(pk_data, aes(x = pred_pop, y = resid, color = subject)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#888", linewidth = 0.8) +
  geom_hline(yintercept = c(-2, 2), linetype = "dotted", color = "#cc4444", linewidth = 0.6) +
  geom_smooth(method = "loess", se = FALSE, aes(group = 1), color = "#FF9800", linewidth = 1) +
  geom_point(size = 2.2, alpha = 0.85) +
  scale_color_manual(values = subject_colors, guide = "none") +
  labs(title = "Residuals vs Predicted", x = "Population Predicted (mg/L)", y = "Normalized Residuals") + pk_theme

p_qq <- ggplot(pk_data, aes(sample = resid)) +
  stat_qq(color = "#2196F3", size = 2.5, alpha = 0.8) +
  stat_qq_line(color = "#E91E63", linewidth = 1, linetype = "dashed") +
  labs(title = "Q-Q Plot of Residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles") + pk_theme

(p_ipred | p_ppred) / (p_restime | p_respred) / (p_qq | plot_spacer())
```

## Individual Fits

```{r ind-fits, fig.height=9}
ggplot(pk_data) +
  geom_line(aes(x = time, y = pred_ind, color = subject), linewidth = 1.1, alpha = 0.9) +
  geom_point(aes(x = time, y = conc, color = subject), size = 2.5,
             shape = 21, fill = "white", stroke = 1.2) +
  facet_wrap(~ subject, ncol = 4, labeller = label_both) +
  scale_color_manual(values = subject_colors, guide = "none") +
  labs(title = "Individual PK Fits",
       subtitle = "Lines = model prediction | Points = observed concentration",
       x = "Time (h)", y = "Concentration (mg/L)") +
  pk_theme
```

---

# Inter-Individual Variability

```{r iiv-plots, fig.height=4}
ranef_df <- ranef(pk_nlme) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("subject") %>%
  rename(lKe = `(Intercept)`) %>%
  tidyr::pivot_longer(-subject, names_to = "parameter", values_to = "eta")

ggplot(ranef_df, aes(x = reorder(subject, eta), y = eta, fill = parameter)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  facet_wrap(~ parameter, scales = "free_y", labeller = label_both) +
  scale_fill_manual(values = c("#2196F3","#4CAF50","#E91E63"), guide = "none") +
  labs(title = "Individual Random Effects (η) by Subject",
       subtitle = "Deviation from population mean",
       x = "Subject", y = "Random Effect (η)") +
  coord_flip() +
  pk_theme
```

---

# Conclusions

::: {.highlight}
1. **Model Performance:** The one-compartment model adequately described theophylline PK in all 12 subjects, with good agreement between observed and predicted concentrations.

2. **Population Parameters:** Mean elimination half-life ≈ 7.9 h; clearance ≈ 0.041 L/h/kg; volume ≈ 0.48 L/kg — consistent with published literature.

3. **IIV:** Moderate inter-individual variability was observed in all three parameters, justifying the mixed-effects approach.

4. **Residual Diagnostics:** Residuals appear symmetrically distributed around zero with no systematic trends, supporting model adequacy.
:::

---

# Session Information

```{r session-info}
sessionInfo()
```

---

*Report generated by the PKPD R Project. See `app.R` for the interactive Shiny dashboard.*
